### 2014-09-19 text/Chapter05.tex
Finish conclusions and tweaks (Chapter 5)

\chapter{Structural coeffect language} 
\label{ch:structural}

@@

\begin{remark}[Let binding]
In a structural coeffect calculus, the typing of $(\lambda x.e_2)~e_1$ is equivalent
to the typing of $\kvd{let}~x=e_1~\kvd{in}~e_2$.
\end{remark}
\begin{proof}
Consider the following typing derivation for $(\lambda x.e_2)~e_1$. Note that 
in the last step, we apply (\emph{exch}) repeatedly to swap $\Gamma_1$ and $\Gamma_2$. 
%
\begin{equation*}
\inference
  { \begin{array}{l}\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1\vspace{-1.6em}\end{array} &
    \inference
      { \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2 }
      { \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash \lambda x.e_2 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2} \vspace{0.2em} }
  { \inference
      { \coctx{\Gamma_2, \Gamma_1}{\aclrd{\textbf{s}} \;\atimes\; (\cclrd{t} \,\aseq\, \aclrd{\textbf{r}})} \vdash (\lambda x.e_2)~e_1 : \tau_2 }
      { \coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\;  \aclrd{\textbf{s}} } \vdash (\lambda x.e_2)~e_1 : \tau_2 } }
\end{equation*}

@@

Similarly to the flat version, each concrete instance of the structural coeffect calculus has a 
different notion of context and thus a different operational interpretation. As before, the 
properties of the flat coeffect algebra guarantee that certain equational properties hold for 
all instances of the calculus.

@@

When discussing equational theory for the flat calculus in Section~\ref{sec:flat-syntax}, we 
noted that no single technique works universally for all flat coeffect calculi. We considered
multiple different reductions that can be used as the basis for operational semantics for
calculi satisfying different additional properties.

@@

\paragraph{Substitution for flat coeffects.} 
The more interesting variant of the substitution lemma for flat coeffects 
(Lemma~\ref{thm:cbn-substitution-bot}) required all operations of the flat coeffect algebra
to coincide. This enables the substitution to preserve the type of expressions, because all
additional requirements arising as the result of the substitution can be associated with the
declaration context. For example, consider the following example where implicit parameter 
$\ident{?offset}$ is substituted for the variable $y$:
%
\begin{equation*}
\begin{array}{rclcll}
 \coctx{y\!:\!\ident{int}}{\cclrd{\emptyset}} &\narrow{\vdash}& \lambda x.y + \ident{?total} &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textnormal{before}) \\
\coctx{()}{\cclrd{ \{ \ident{?offset} \} }} &\narrow{\vdash}& \lambda x.\ident{?offset} + \ident{?total} &\narrow{:}&
  \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textnormal{after})
\end{array}  
\end{equation*}
%
The typing judgement obtained in (\emph{after}) preserves the type of the expression
(function value) from the original typing (\emph{before}). This is possible thanks to the
non-determinism involved in lambda abstraction -- as all operators of the flat coeffect
algebra used here are $\cup$, we can place the additional requirement on the outer context.
Note that this is not the \emph{only} possible typing, but it is \emph{permissible} typing.

@@

\paragraph{Substitution for structural coeffects.}
In contrast, the substitution lemma for structural coeffects can be proven ecause 
structural coeffect systems provide enough \emph{precision} to identify exactly with 
which variable should a context requirement be associated.

The following example shows a situation similar to the previous one -- here, we use
structural data-flow calculus (with the \kvd{prev} construct to obtain previous
value of an expression) and we substitute $w+z$ for $y$:
%
\begin{equation*}
\begin{array}{rclcll}
 \coctx{y\!:\!\ident{int}}{ \alift{\cclrd{2}} } 
  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~y) &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textnormal{before}) \\
\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\cclrd{2}\cclrd{\ast}\alift{\cclrd{1},\cclrd{1}}} 
  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textnormal{after}) \\
\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\alift{\cclrd{2},\cclrd{2}}} 
  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textnormal{final})
\end{array}  
\end{equation*}
%
The type of the function does not change, because the structural type system associates
the annotation $\cclrd{1}$ with the bound variable $x$ and the substitution does not 
affect how the variable $x$ is used.

@@

As demonstrated in the previous section, reduction (and substitution) in the structural coeffect 
calculus may need to replace a \emph{single} variable with a \emph{vector} of variables. 
Furthermore, we may also need to substitute for multiple variables in the variable context.

For example, the expression $\lambda x.x+x$ is type-checked by type-checking $x_1 + x_2$,
contracting $x_1$ and $x_2$ and then applying lambda abstraction. During the reduction 
$(\lambda x.x+x)~(y+z)$ we thus need to substitute $y_1+z_1$ for $x_1$ and $y_2+z_2$ for $x_2$.
This is similar to substitution lemma in other structural variants of $\lambda$-calculus,
for example in the bunched typing system \cite{substruct-bunched}.
To express the substitution lemma, we define the notion of a \emph{context with holes}

@@

\noindent
A context with $n$ holes may either start with a hole, followed by a context with $n-1$ holes, or it may
start with a variable followed by a context with $n$ holes. Note that the definition ensures that
the location of variable holes correspond to the locations of coeffect annotation holes. Given a context
with holes, we can fill the holes with other contexts using the \emph{hole filling} operation and
obtain an ordinary coeffect-annotated context.

@@

Substitution can be seen as sequential composition. Informally -- we first need to obtain the 
value of the expression (requiring $\aclrd{\textbf{t}}$) and then use it in context with 
requirements $\cclrd{s}$. Thus the free variables of the expression \emph{after} substitution
are annotated with $\cclrd{s} \aseq \aclrd{\textbf{t}}$, using the (scalar-vector extension)
of the sequential composition operator $\cseq$.

\begin{lemma}[Multi-nary substitution]
Given an expression with multiple holes filled by variables $x_i\!:\!\tau_i$ with coeffects $\cclrd{s_k}$:
%
\begin{equation*}
\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{x_1\!:\!\tau_1}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
  \coctx{x_k\!:\!\tau_k}{\alift{\cclrd{s_k}}}\,] \vdash e_r : \tau_r
\end{equation*}
%
and a expressions $e_i$ with free-variable contexts $\Gamma_i$ annotated with $\aclrd{\mathbf{t_i}}$:
%
\begin{equation*}
\coctx{\Gamma_1}{\aclrd{\textbf{t}_1}} \vdash e_1 : \tau_1 
\quad \ldots \quad
\coctx{\Gamma_k}{\aclrd{\textbf{t}_k}} \vdash e_k : \tau_k
\end{equation*}
%
then substituting the expressions $e_i$ for variables $x_i$ results in an expression with a context
where the original holes are filled by contexts $\Gamma_i$ with coeffects $\cclrd{s_i} \aseq \aclrd{\mathbf{t_i}}$:
%
\begin{equation*}
\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{\Gamma_s}{\cclrd{s_1}\aseq\,\aclrd{\textbf{t}_1}} \,|\, \ldots \,|\, 
  \coctx{\Gamma_s}{\cclrd{s_k}\aseq\,\aclrd{\textbf{t}_k}}\,] \vdash \subst{\subst{e_r}{x_1}{e_1}\ldots}{x_k}{e_k} : \tau_r
\end{equation*}
\end{lemma}
\begin{proof}
By induction over $\vdash$, using the multi-nary aspect of the substitution 
in the proof of the contraction case. See Appendix~?.
\end{proof}

@@

\todo{Exterminate!}

~

Because of the vector (free monoid) structure, coeffects $\aclrd{R_1}$, $\aclrd{R_2}$, and 
$\langle{\cclrd{r}}\rangle$ for the receiving term $e_r$ are uniquely associated with $\Gamma_1$, $\Gamma_2$, and
$x$ respectively. Therefore, substituting $e_s$ (which has coeffects $S$) for $x$ introduces
the context dependencies specified by $S$ which are composed with the requirements $r$ on $x$. 
Using the substitution lemma, we can demonstrate $\beta$-equality:
%
\begin{equation*}
\inference{
  \inference{\coctx{\Gamma_1, x : \sigma}{\aclrd{R} \cons \langle{\cclrd{r}}\rangle} \vdash e_1 : \tau}
             {\coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{r}} \tau} &
           \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
  {\coctx{\Gamma_1,\Gamma_2}{\aclrd{R} \cons (\cclrd{r} \,\aseq\, \aclrd{S})} \vdash (\lambda x . e_1) e_2 
\equiv
\subst{e_1}{x}{e_2} : \tau}
\end{equation*}
%%
\noindent
As a result, $\beta$-reduction preserves the type and coeffects of a term.
This gives the following subject reduction property:

\begin{theorem}[Subject reduction]
In a structural coeffect calculus, if $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$ and
$e \reduce_\beta e'$ then $\coctx{\Gamma}{\aclrd{R}} \vdash e' : \tau$.
\end{theorem}
\begin{proof}
Following from Lemma~\ref{thm:structural-subst} and $\beta$-equality.  
\end{proof}

\noindent
Structural coeffect systems also exhibit $\eta$-equality, therefore satisfying both
\emph{local soundness} and \emph{local completeness} conditions set by Pfenning and Davies
\cite{logic-modal-reconstruction}. This means that abstraction does not introduce too much, 
and application does not eliminate too much.
%
\begin{equation*}
\inference{
  \inference 
    {\coctx{\Gamma}{\aclrd{R}} \vdash e : \sigma \xrightarrow{s} \tau & 
     \coctx{x:\sigma}{\alift{\cunit}} \vdash x : \sigma}
    {\coctx{\Gamma, x:\sigma}{\aclrd{R} \,\cons\, (\cclrd{s} \,\aseq\, \alift{\cunit})} \vdash e \, x : \tau}}
  {\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x.e~x  \equiv e 
: \sigma \xrightarrow{\cclrd{s}} \tau}
\end{equation*}
%
The last step uses the fact that 
$\cclrd{s} \,\aseq\, \alift{\cunit} = \alift{\cclrd{s} \,\cseq\, \cunit} = \alift{\cclrd{s}}$ 
arising from the monoid $(\C, \cseq, \cunit)$ of the scalar coeffect structure.

This highlights another difference between coeffects and effects, as $\eta$-equality does
not hold for many notions of effect. For example, in a language with output
effects, $e = (\textsf{print \texttt{"hi"}}; (\lambda x . x))$ has different effects to 
its $\eta$-converted form $\lambda x . e x$ because the immediate effects of $e$ are
hidden by the purity of $\lambda$-abstraction. In the coeffect calculus, the (abs) rule
allows immediate contextual requirements of $e$ to ``float outside'' of the enclosing
$\lambda$. Furthermore, the free monoid nature of $\cons$ in structural systems allows 
the exact immediate requirements of $\lambda x . e x$ to match those of $e$.

\newpage

@@

\todo{ (...) }

>>>>

\chapter{Structural coeffect calculus} 
\label{ch:structural}

@@

\begin{remark}[Let binding]
\label{thm:structural-letbinding}
In a structural coeffect calculus, the typing of $(\lambda x.e_2)~e_1$ is equivalent
to the typing of $\kvd{let}~x=e_1~\kvd{in}~e_2$.
\end{remark}
\begin{proof}
Consider the following typing derivation for $(\lambda x.e_2)~e_1$. Note that 
in the last step, we apply (\emph{exch}) repeatedly to swap $\Gamma_1$ and $\Gamma_2$. 
%
\begin{equation*}
\inference
  { \begin{array}{l}\coctx{\Gamma_1}{\aclrd{\textbf{r}}} \vdash e_1 : \tau_1\vspace{-1.6em}\end{array} &
    \inference
      { \coctx{\Gamma_2, x\!:\!\tau_1}{\aclrd{\textbf{s}}\;\atimes\;\alift{\cclrd{t}}} \vdash e_2 : \tau_2 }
      { \coctx{\Gamma_2}{\aclrd{\textbf{s}}} \vdash \lambda x.e_2 : \tau_1 \xrightarrow{\cclrd{t}} \tau_2} \vspace{0.2em} }
  { \inference
      { \coctx{\Gamma_2, \Gamma_1}{\aclrd{\textbf{s}} \;\atimes\; (\cclrd{t} \,\aseq\, \aclrd{\textbf{r}})} \vdash (\lambda x.e_2)~e_1 : \tau_2 }
      { \coctx{\Gamma_1, \Gamma_2}{(\cclrd{t} \,\aseq\, \aclrd{\textbf{r}}) \;\atimes\;  \aclrd{\textbf{s}} } \vdash (\lambda x.e_2)~e_1 : \tau_2 } }
\end{equation*}

@@

Similarly to the flat version, each concrete instance of the structural coeffect calculus has a 
different notion of context and thus a different operational interpretation. As before, the 
properties of the flat coeffect algebra guarantee that certain equational properties hold for 
all instances of the calculus. In this section, we look at these common properties that we get
``for free'' in all structural coeffect calculi.

@@

When discussing equational theory for the flat calculus in Section~\ref{sec:flat-syntax}, we 
noted that no single technique of specifying syntactic reduction works universally for all 
flat coeffect calculi. We considered multiple different options (call-by-name, call-by-value, 
internalized substitution) that can be used as the basis for operational semantics for
different calculi that satisfy different additional properties.

@@

\paragraph{Substitution for flat coeffects.} 
The less obvious (\emph{top-pointed}) variant of the substitution lemma for flat coeffects 
(Lemma~\ref{thm:cbn-substitution-bot}) required all operations of the flat coeffect algebra
to coincide. This enables the substitution to preserve the type of expressions, because all
additional requirements arising as the result of the substitution can be associated with the
declaration context. For example, consider the following example where implicit parameter 
$\ident{?offset}$ is substituted for the variable $y$:
%
\begin{equation*}
\begin{array}{rclcll}
 \coctx{y\!:\!\ident{int}}{\cclrd{\emptyset}} &\narrow{\vdash}& \lambda x.y + \ident{?total} &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textnormal{before}) \\
\coctx{()}{\cclrd{ \{ \ident{?offset} \} }} &\narrow{\vdash}& \lambda x.\ident{?offset} + \ident{?total} &\narrow{:}&
  \ident{int} \xrightarrow{\cclrd{ \{ \ident{?total} \} }} \ident{int} & \qquad(\textnormal{after})
\end{array}  
\end{equation*}
%
The typing judgement obtained in (\emph{after}) preserves the type of the expression
(function value) from the original typing (\emph{before}). This is possible thanks to the
non-determinism involved in lambda abstraction -- as all operators of the flat coeffect
algebra used here are $\cup$, we can place the additional requirement on the outer context.
Note that this is not the \emph{only} possible typing, but it is \emph{permissible} typing.

@@

\paragraph{Substitution for structural coeffects.}
In contrast, the substitution lemma for structural coeffects can be proven because 
structural coeffect systems provide enough \emph{precision} to identify exactly with 
which variable should a context requirement be associated.

The following example shows a situation similar to the previous one -- here, we use
structural data-flow calculus (writing $\kvd{prev}~e$ to obtain previous value of the 
expression $e$) and we substitute $w+z$ for $y$:
%
\begin{equation*}
\begin{array}{rclcll}
 \coctx{y\!:\!\ident{int}}{ \alift{\cclrd{2}} } 
  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~y) &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textnormal{before}) \\
\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\cclrd{2}\cclrd{\ast}\alift{\cclrd{1},\cclrd{1}}} 
  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textnormal{after}) \\
\coctx{w\!:\!\ident{int}, z\!:\!\ident{int}}{\alift{\cclrd{2},\cclrd{2}}} 
  &\narrow{\vdash}& \lambda x.\kvd{prev}~(x + \kvd{prev}~(w+z))) &\narrow{:}&
   \ident{int} \xrightarrow{\cclrd{ 1 }} \ident{int} & \quad(\textnormal{final})
\end{array}  
\end{equation*}
%
The type of the function does not change, because the structural type system associates
the annotation $\cclrd{1}$ with the bound variable $x$ and the substitution does not 
affect how the variable $x$ is used.

@@

As demonstrated in the previous section, reduction (and substitution) in the structural coeffect 
calculus may need to replace a \emph{single} variable with a \emph{vector} of variables. 
More importantly, because the system uses explicit contraction, we may also need to substitute 
for multiple variables in the variable context at the same time.

Consider the expression $\lambda x.x+x$. It is type-checked by type-checking $x_1 + x_2$,
contracting $x_1$ and $x_2$ and then applying lambda abstraction. During the reduction of 
$(\lambda x.x+x)~(y+z)$ we need to substitute $y_1+z_1$ for $x_1$ and $y_2+z_2$ for $x_2$.
This is similar to substitution lemma in other structural variants of $\lambda$-calculus,
such as the bunched typing system \cite{substruct-bunched}. To express the substitution 
lemma, we define the notion of a \emph{context with holes}:

@@

\noindent
A context with $n$ holes may either start with a hole, followed by a context with $n-1$ holes, or it may
start with a variable followed by a context with $n$ holes. Note that the definition ensures that
the locations of variable holes correspond to the locations of coeffect annotation holes. Given a context
with holes, we can fill the holes with other contexts using the \emph{hole filling} operation and
obtain an ordinary coeffect-annotated context.

@@

Substitution can be seen as sequential composition. Informally -- we first need to obtain the 
value of the expression (requiring $\aclrd{\textbf{t}}$) and then use it in context with 
requirements $\cclrd{s}$. Thus the free variables of the expression \emph{after} substitution
are annotated with $\cclrd{s} \,\aseq\, \aclrd{\textbf{t}}$, using the (scalar-vector extension)
of the sequential composition operator $\cseq$. This suggests that, to evaluate each of the variables
annotated with coeffects in the vector $\aclrd{\textbf{t}}$, we first need to evaluate the substituted
expression with coeffects $\cclrd{s}$, followed by the rest of the expression with coeffects specified
by $\aclrd{\textbf{t}}$.

\begin{lemma}[Multi-nary substitution]
\label{thm:structural-substitution}
Given an expression with multiple holes filled by variables $x_i\!:\!\tau_i$ with coeffects $\cclrd{s_k}$:
%
\begin{equation*}
\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{x_1\!:\!\tau_1}{\alift{\cclrd{s_1}}} \,|\, \ldots \,|\,
  \coctx{x_k\!:\!\tau_k}{\alift{\cclrd{s_k}}}\,] \vdash e_r : \tau_r
\end{equation*}
%
and a expressions $e_i$ with free-variable contexts $\Gamma_i$ annotated with $\aclrd{\mathbf{t_i}}$:
%
\begin{equation*}
\coctx{\Gamma_1}{\aclrd{\textbf{t}_1}} \vdash e_1 : \tau_1 
\quad \ldots \quad
\coctx{\Gamma_k}{\aclrd{\textbf{t}_k}} \vdash e_k : \tau_k
\end{equation*}
%
then substituting the expressions $e_i$ for variables $x_i$ results in an expression with a context
where the original holes are filled by contexts $\Gamma_i$ with coeffects $\cclrd{s_i} \aseq \aclrd{\mathbf{t_i}}$:
%
\begin{equation*}
\coctx{\Gamma}{\aclrd{\textbf{r}}}~[\,\coctx{\Gamma_s}{\cclrd{s_1}\aseq\,\aclrd{\textbf{t}_1}} \,|\, \ldots \,|\, 
  \coctx{\Gamma_s}{\cclrd{s_k}\aseq\,\aclrd{\textbf{t}_k}}\,] \vdash \subst{\subst{e_r}{x_1}{e_1}\ldots}{x_k}{e_k} : \tau_r
\end{equation*}
\end{lemma}
\begin{proof}
By induction over $\vdash$, using the multi-nary aspect of the substitution 
in the proof of the contraction case. See Appendix~?.
\end{proof}

@@

In the previous chapter, we discussed call-by-value separately from call-by-name, because the
proof of call-by-value substitution has fewer prerequisites. In this section, we consider full
$\beta$-reduction (local soundness), which encompasses both types of evaluation and 
$\eta$-expansion (local completeness). Both of the properties hold for a system with any
structural coeffect algebra.

\paragraph{Reduction theorem}
In a full $\beta$-reduction,
written as $\rightarrow_\beta$, we replace the redex $(\lambda x.e_2)~e_1$ by the
expression $\subst{e_r}{x}{e_s}$ anywhere inside a term. The subject reduction theorem
guarantees that this does not change the type of the term.

\begin{theorem}[$\beta$-reduction]
In a structural coeffect system with a structural coeffect algebra formed by 
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and operations $\alift{-}$ and $\aseq$,
if $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ 
and $e \rightarrow_{\beta} e'$ using the full $\beta$-reduction then
$\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e' : \tau$.
\end{theorem}
\begin{proof}
Consider the typing derivation for the redex $(\lambda x.e_r)\,e_s$ before the reduction
(note that this is similar to the typing of let binding discussed in Remark~\ref{thm:structural-letbinding},
but we do not swap the two parts of the free-variable context):
%
\begin{equation*}
\inference
  { \begin{array}{l}\coctx{\Gamma_s}{\aclrd{\textbf{s}}} \vdash e_s : \tau_s\vspace{-2em}\end{array} &
    \inference
      { \coctx{\Gamma_r, x\!:\!\tau_s}{\aclrd{\textbf{r}}\;\atimes\,\alift{\cclrd{t}}} \vdash e_r : \tau_r }
      { \coctx{\Gamma_r}{\aclrd{\textbf{r}}} \vdash \lambda x.e_r : \tau_s \xrightarrow{\cclrd{t}} \tau_r} \vspace{0.2em} }
  { \coctx{\Gamma_r, \Gamma_s}{\aclrd{\textbf{r}} \;\atimes\; (\cclrd{t} \,\aseq\, \aclrd{\textbf{s}})} \vdash (\lambda x.e_r)~e_s : \tau_r }
\end{equation*}
%
For the substitution lemma, we first rewrite the typing judgement for $e_r$, \ie~
$\coctx{\Gamma_r, x\!:\!\tau_s}{\aclrd{\textbf{r}}\;\atimes\,\alift{\cclrd{t}}} \vdash e_r : \tau_r$
as a context with a single hole filled by the $x$ variable:
$\coctx{\Gamma_r, -}{\aclrd{\textbf{r}}\;\atimes\,-}~[\,\coctx{x\!:\!\tau_s}{\alift{\cclrd{t}}}\,] \vdash e_r : \tau_r$.
Now we can perform the substitution using Lemma~\ref{thm:structural-substitution}:
%
\begin{equation*}
\inference
  { \coctx{\Gamma_r, -}{\aclrd{\textbf{r}}\;\atimes-}~[\,\coctx{x\!:\!\tau_s}{\alift{\cclrd{t}}}\,] \vdash e_r : \tau_r & 
    \coctx{\Gamma_s}{\aclrd{\textbf{s}}} \vdash e_s : \tau_s }
  { \inference
      { \coctx{\Gamma_r, -}{\aclrd{\textbf{r}}\;\atimes-}~[\,\coctx{\Gamma_s}{\cclrd{t}\,\aseq\,\aclrd{\textbf{s}} }\,] \vdash \subst{e_r}{x}{e_s} : \tau_r }
      { \coctx{\Gamma_r, \Gamma_s}{\aclrd{\textbf{r}}\,\atimes\,(\cclrd{t}\,\aseq\,\aclrd{\textbf{s}})} \vdash \subst{e_r}{x}{e_s} : \tau_r } }
\end{equation*}
%
The last step applies the hole filling operation, showing that substitution preserves 
the type of the term.
\end{proof}

\noindent
Because of the vector (free monoid) structure, coeffect annotations $\aclrd{\textbf{r}}$, 
$\aclrd{\textbf{s}}$, and $\alift{\cclrd{t}}$ are uniquely associated with $\Gamma_r$, 
$\Gamma_s$, and $x$ respectively. Therefore, substituting $e_s$ (which has coeffects 
$\aclrd{\textbf{s}}$) for $x$ introduces the context dependencies specified by $\aclrd{\textbf{s}}$
which are composed with the existing requirements $\cclrd{t}$ on $x$.

\paragraph{Expansion theorem}
Structural coeffect systems also exhibit $\eta$-equality, therefore satisfying both
\emph{local soundness} and \emph{local completeness}. Informally, this means that 
abstraction does not introduce too much, and application does not eliminate too much.

\begin{theorem}[$\eta$-expansion]
In a structural coeffect system with a structural coeffect algebra formed by 
$(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ and operations $\alift{-}$ and $\aseq$,
if $\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e : \tau$ 
and $e \rightarrow_{\eta} e'$ using the full $\eta$-reduction then
$\coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash e' : \tau$.
\end{theorem}
\begin{proof}
The following derivation shows that $\lambda x.f~x$ has the same type as $f$:
\begin{equation*}
\inference
  { \inference
    { \coctx{\Gamma}{\aclrd{\textbf{r}}} \vdash f : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 &
      \coctx{x:\tau_1}{\alift{\cunit}} \vdash x:\tau_1 }
    { \coctx{\Gamma, x:\tau_1}{\aclrd{\textbf{r}} \;\atimes\;(\cclrd{s}\,\aseq\,\alift{\cunit}) } \vdash f~x : \tau_2 } }
  { \inference
    { \coctx{\Gamma, x:\tau_1}{\aclrd{\textbf{r}} \;\atimes\,\alift{\cclrd{s}} } \vdash f~x : \tau_2 }
    { \coctx{\Gamma}{ \aclrd{\textbf{r}} } \vdash \lambda x.f~x : \tau_1 \xrightarrow{\cclrd{s}} \tau_2 } }
\end{equation*}
%
The second step uses the fact that 
$\cclrd{s} \,\aseq\, \alift{\cunit} = \alift{\cclrd{s} \,\cseq\, \cunit} = \alift{\cclrd{s}}$ 
arising from the monoid $(\C, \cseq, \cunit)$ of the scalar coeffect structure.
\end{proof}

\noindent
The $\eta$-expansion property discussed in this section highlights another difference 
between coeffects and effects. The $\eta$-equality property does not hold for many notions 
of effect. For example, in a language with output effects, $e = (\textsf{print \texttt{"hi"}}; (\lambda x . x))$ 
has different effects to its $\eta$-converted form $\lambda x . e x$ because the immediate 
effects of $e$ are hidden by the purity of $\lambda$-abstraction. In the coeffect calculus, 
the (\emph{abs}) rule allows immediate contextual requirements of $e$ to ``float outside'' 
of the enclosing $\lambda$. Furthermore, the free monoid nature of $\atimes$ in structural 
coeffect systems allows the exact immediate requirements of $\lambda x . e x$ to match 
those of $e$.

% ==================================================================================================
%
%     ###                         ##                    #                 
%    #   #                         #                                      
%    #       ###   # ##    ###     #    #   #   ###    ##     ###   # ##  
%    #      #   #  ##  #  #   #    #    #   #  #        #    #   #  ##  # 
%    #      #   #  #   #  #        #    #   #   ###     #    #   #  #   # 
%    #   #  #   #  #   #  #   #    #    #  ##      #    #    #   #  #   # 
%     ###    ###   #   #   ###    ###    ## #  ####    ###    ###   #   # 
%
% ==================================================================================================

@@

This chapter completes the key development of this thesis -- the presentation of unified calculi
for context-aware computations that capture the motivating examples introduced in 
Chapter~\ref{ch:applications}. In the previous chapter, we focused on the first category of 
contet-aware computations and we developed \emph{flat coeffect calculus} that captures
\emph{whole-context} properties. This chapter develops \emph{structural coeffect calculus},
capturing \emph{per-variable} contextual properties. The system provides a precise 
analysis of liveness and data-flow and allows other interesting uses such as tracking of variable
accesses based on bounded linear logic.

Following the structure of the previous chapter, the structural coeffect calculus is parameterized
by a \emph{structural coeffect algebra}. The two definitions are similar -- both require operations 
$\cseq$ and $\cpar$ that model sequential and point-wise composition, respectively. For flat 
coeffects, we further require $\czip$ to model context merging. For structural coeffects, we 
instead use a vector (free monoid) with the $\atimes$ operation -- which serves a similar purpose
as $\czip$.

In order to keep track of separate annotations for each variable, we use a system with 
explicit structural rules (contraction, weakening and exchange) that manipulate the structure
of variables and the structure of annotations at the same time.

The chapter presents two technical results. The first is the semantics of structural coeffect
calculus in terms of \emph{structural indexed comonads}. The semantics serves two purposes --
it unifies our three example calculi (bounded reuse, data-flow and liveness) and it demonstrates
suitability of the type system. The second technical result is equational theory for structural
coeffect calculi. In particular, we show that $\beta$-reduction and $\eta$-expansion preserve
the typing for any structural coeffect calculi. These two strong properties are desirable for
programming languages, but are often not satisfied (\eg~by effectful langauges).

Two questions remain opened for the upcoming chapter. Firstly, is there a way to present 
flat and structural coeffect calculi in a uniform way as a single system? Secondly, can we
simplify the definitions to make type inference easier and add other practical language
aspects such as recursion?
